5. Cluster Evaluation

### 2024-11-11 tutorial/05-evaluation.qmd
tutorial update
config updated

lib <- c(
  "rmarkdown",
  "knitr",
  "tidyverse",
  "kableExtra",
  "tsFeatureExtracR",
  "factoextra",
  "fpc"
)
invisible(lapply(lib, library, character.only = TRUE))
rm(lib)

@@

set.seed(42)
kmeans_final <- kmeans(pca_scores, centers = 2, nstart = 100)

@@

```{r}
#| label: Extract cluster assignments and summarize
#| message: false
#| warning: false

featCluster <- data.frame(ID = as.numeric(names(kmeans_final$cluster)), 
                            cluster = kmeans_final$cluster) %>%
  left_join(z_data, by = "ID") %>%
  left_join(
    featData %>% 
      select(ID, EvDayDiscr.post) %>% 
      distinct(ID, EvDayDiscr.post, .keep_all = TRUE) %>%
      mutate(EvDayDiscr.post_z = scale(EvDayDiscr.post, scale = TRUE)) %>%
      replace_na(list(EvDayDiscr.post_z = NA)) %>%
      select(
        ID,
        EvDayDiscr.post_z
      ), 
    by = "ID"
  ) %>%
  select(-ID) %>%
  group_by(cluster) %>%
  summarise(across(everything(), list(mean = mean), na.rm=TRUE)) %>%
  ungroup %>% 
  select(-starts_with("ID")) %>%
  pivot_longer(
    cols = -c(cluster),
    names_to = c("variable", ".value"),
    names_pattern = "(.*)_(.*)"
  ) %>% 
  mutate(
    nam = variable,
    variable = gsub("\\_.*", "", nam),
    feature = gsub("^.*?\\_", "", nam)
  ) %>% 
  mutate(feature = ifelse(feature == "z", "mean", feature)) %>%
  left_join(var_meta %>% select(name, label), by = c("variable" = "name")) %>%
  select(
    cluster,
    nam,
    variable,
    label,
    feature,
    everything()
  )
```

@@

# clusterVariableGrid
clusterVariableGrid <- featCluster %>% 
  filter(variable != "EvDayDiscr.post",
         feature != "n") %>%
  ggplot(., aes(x = feature, y = mean, group = cluster, color = as.factor(cluster))) +
  geom_point() +
  geom_line() +
  scale_colour_manual(values = c("#384B6B", "#E2892B")) + 
  #geom_errorbar(aes(ymin = mean-se, ymax = mean+se), width=.1, position=position_dodge(0.1), size=0.5, alpha=0.4) +
  coord_flip() +
  labs(
    x = "Features",
    y = "Standardized Mean",
    color = "Cluster"
  ) +
  facet_wrap(~ label) +
  theme_Publication()

# clusterFeatureGrid
clusterFeatureGrid <- featCluster %>% 
  ggplot(., aes(x = reorder(label, -as.numeric(label == "Post: Discrimination")), y = mean, group = cluster, color = as.factor(cluster))) +
  geom_point() +
  geom_line() +
  #geom_ribbon(aes(ymin = mean-se, ymax = mean+se), alpha = 0.2) +
  #geom_errorbar(aes(ymin = mean-se, ymax = mean+se), width=.1, position=position_dodge(0.1), size=0.5, alpha=0.4) +
  scale_colour_manual(values = c("#384B6B", "#E2892B")) + 
  coord_flip() +
  labs(
    x = "Variable",
    y = "Standardized Mean",
    color = "Cluster"
  ) +
  facet_wrap(~ factor(feature, levels = unique(featCluster$feature)), ncol=4) +
  theme_Publication()

ggpubr::ggarrange(
  clusterVariableGrid + theme(axis.title.x = element_blank()),
  clusterFeatureGrid,
  ncol = 1,
  labels = c("(A) Variable Focus", "(B) Feature Focus"),
  common.legend = TRUE,
  legend = "bottom",
  align = "v"
) 
```

@@

clusterTSTrend <- rawCluster %>% 
  ggplot(., aes(x=TIDnum, y=value, group=cluster, color=as.factor(cluster))) +
  geom_line(aes(x=TIDnum, y=value, group=ID, color=as.factor(cluster)), alpha=0.05) +
  geom_smooth(aes(group=cluster)) +
  scale_colour_manual(values = c("#384B6B", "#E2892B")) + 
  #stat_summary(fun=mean, geom="line") +
  facet_wrap(~ label) +
  labs(
    x = "Time ID",
    y = "Response (Mean Cluster Response Bold)",
    color = "Cluster"
  ) +
  theme_Publication()

ggpubr::ggarrange(
  clusterTS + theme(axis.title.x = element_blank()),
  clusterTSTrend,
  ncol = 1,
  labels = c("(A) Timepoint Means", "(B) Average Spline Trends"),
  common.legend = TRUE,
  legend = "bottom"
)
```

@@

### Out-of-cluster Comparison

@@

To illustrate artifact checks, we added the number of ESM measurements into the comparison and find that the participants in the deterioration cluster (cluster 2) on average completed slightly more ESM surveys in general and reported on more intergroup interactions in particular (see $n$ in [@fig-clusterFeatVarComb] B). This difference in the reported number of interactions might indicate either a clustering artifact or a meaningful difference. The higher average number of interactions in cluster 2 could, for example, indicate a clustering artifact if variances are substantially larger due to the larger samples (e.g., restriction of range in the smaller sample; @kogan2006). In our case, this seems less likely because one out of four variables did not differ in terms of the MAD (i.e., our selected measurement of the time series variance; see [@fig-mad-cluster-comparison] below).

```{r}
#| label: fig-mad-cluster-comparison
#| message: false
#| warning: false
#| fig-cap: Cluster comparison MAD across variables

@@

# validation feature
validation_feature <-
  featData %>% 
  select(ID, EvDayDiscr.post) %>% 
  distinct(ID, EvDayDiscr.post, .keep_all = TRUE) %>%
  mutate(EvDayDiscr.post_z = scale(EvDayDiscr.post, scale = TRUE)) %>%
  replace_na(list(EvDayDiscr.post_z = NA)) %>%
  select(
    ID,
    EvDayDiscr.post_z
  ) %>%
  left_join(data.frame(ID = as.numeric(names(kmeans_final$cluster)),
             cluster = kmeans_final$cluster), by = "ID")

t_discrimination <- t.test(
  validation_feature$EvDayDiscr.post_z[validation_feature$cluster == 1],
  validation_feature$EvDayDiscr.post_z[validation_feature$cluster == 2],
  na.action = na.omit
)

results_df <- rbind(
  results_df,
  data.frame(
    concept = "EvDayDiscr.post",
    feature = "mean",
    difference = t_discrimination$estimate[1] - t_discrimination$estimate[2],
    t_value = t_discrimination$statistic,
    df = t_discrimination$parameter,
    p_value = t_discrimination$p.value,
    lwr = t_discrimination$conf.int[1],
    upr = t_discrimination$conf.int[2]
  )
) %>%
  annotate_significance(., 
                        variableLab = var_meta %>% select(name, label), 
                        join_by = c('concept' = 'name'))

feature_plot_list <- generate_feature_plots(results_df)

feature_plot_list$mad
```

>>>>

lib <- c(
  "rmarkdown",
  "knitr",
  "tidyverse",
  "kableExtra",
  "tsFeatureExtracR",
  "factoextra",
  "fpc",
  "shiny",
  "shinyWidgets",
  "shinyjs",
  "bslib"
)
invisible(lapply(lib, library, character.only = TRUE))
rm(lib)

@@

set.seed(12345)
kmeans_final <- kmeans(pca_scores, centers = 2, nstart = 100)
kmeans_final$cluster

saveRDS(kmeans_final, file = "data/kmeans_final.Rds")
```

@@

::: scrolling
```{r}
#| label: Extract cluster assignments and summarize
#| message: false
#| warning: false

# calculate number of removed datapoints
missingness_ooc_id <- full_join(
  featData %>%
    mutate(uid = paste(study, PID, sep = "_")) %>%
    group_by(uid, ID) %>%
    summarise(n_feature = n()) %>%
    ungroup(),
  dt_raw %>%
    mutate(uid = paste(study, PID, sep = "_")) %>%
    group_by(uid) %>%
    summarise(n_raw = n()) %>%
    ungroup(),
  by = "uid"
) %>%
  mutate(
    n_rm = n_raw - n_feature
  ) %>%
  filter(
    !is.na(ID)
  ) %>%
  mutate(nrm_z = scale(n_rm)) %>%
  select(ID, nrm_z)
discrimination_ooc_id <- featData %>% 
  select(ID, EvDayDiscr.post) %>% 
  distinct(ID, EvDayDiscr.post, .keep_all = TRUE) %>%
  mutate(EvDayDiscr.post_z = scale(EvDayDiscr.post, scale = TRUE)) %>%
  replace_na(list(EvDayDiscr.post_z = NA)) %>%
  select(
    ID,
    EvDayDiscr.post_z
  )

featCluster <- data.frame(ID = as.numeric(names(kmeans_final$cluster)), 
                            cluster = kmeans_final$cluster) %>%
  left_join(z_data, by = "ID") %>%
  left_join(discrimination_ooc_id, by = "ID") %>% 
  left_join(missingness_ooc_id, by = "ID") %>%
  select(-ID) %>%
  group_by(cluster) %>%
  summarise(across(everything(), list(mean = mean), na.rm=TRUE)) %>%
  ungroup %>% 
  select(-starts_with("ID")) %>%
  pivot_longer(
    cols = -c(cluster),
    names_to = c("variable", ".value"),
    names_pattern = "(.*)_(.*)"
  ) %>% 
  mutate(
    nam = variable,
    variable = gsub("\\_.*", "", nam),
    feature = gsub("^.*?\\_", "", nam)
  ) %>% 
  mutate(feature = ifelse(feature == "z", "mean", feature)) %>%
  left_join(var_meta %>% select(name, label) %>% rbind(c("nrm", "OOC: Measurements removed")), by = c("variable" = "name")) %>%
  select(
    cluster,
    nam,
    variable,
    label,
    feature,
    everything()
  ) %>% 
  mutate(feature = gsub("^n$", "n (within ooc)", feature)) %>% 
  mutate(feature = gsub("^mean$", "between ooc (mean)", feature))

```
:::

@@

# clusterVariableGrid
clusterVariableGrid <- featCluster %>% 
  filter(variable != "EvDayDiscr.post",
         variable != "nrm",
         feature != "n (within ooc)") %>%
  ggplot(., aes(x = feature, y = mean, group = cluster, color = as.factor(cluster))) +
  geom_point() +
  geom_line() +
  scale_colour_manual(values = c("#384B6B", "#E2892B")) + 
  #geom_errorbar(aes(ymin = mean-se, ymax = mean+se), width=.1, position=position_dodge(0.1), size=0.5, alpha=0.4) +
  coord_flip() +
  labs(
    x = "Features",
    y = "Standardized Mean",
    color = "Cluster"
  ) +
  facet_wrap(~ label) +
  theme_Publication()

# clusterFeatureGrid
clusterFeatureGrid <- featCluster %>% 
  ggplot(., aes(x = reorder(label, as.numeric(!label %in% c("OOC: Discrimination", "OOC: Measurements removed"))), y = mean, group = cluster, color = as.factor(cluster))) +
  geom_point() +
  geom_line() +
  #geom_ribbon(aes(ymin = mean-se, ymax = mean+se), alpha = 0.2) +
  #geom_errorbar(aes(ymin = mean-se, ymax = mean+se), width=.1, position=position_dodge(0.1), size=0.5, alpha=0.4) +
  scale_colour_manual(values = c("#384B6B", "#E2892B")) + 
  coord_flip() +
  labs(
    x = "Variable",
    y = "Standardized Mean",
    color = "Cluster"
  ) +
  facet_wrap(~ factor(feature, levels = unique(featCluster$feature)), ncol=4) +
  theme_Publication()

clusterFeatVarComb <- ggpubr::ggarrange(
  clusterVariableGrid + theme(axis.title.x = element_blank()),
  clusterFeatureGrid,
  ncol = 1,
  labels = c("(A) Variable Focus", "(B) Feature Focus"),
  common.legend = TRUE,
  legend = "bottom",
  align = "v"
) 
ggsave("../Figures/clusterFeatVarComb_tutorial.pdf", clusterFeatVarComb, width = 12, height = 12)
clusterFeatVarComb
```

@@

clusterTSTrend <- rawCluster %>% 
  ggplot(., aes(x=TIDnum, y=value, group=cluster, color=as.factor(cluster))) +
  geom_line(stat="smooth", method = "gam", aes(group=ID, color=as.factor(cluster)), se = FALSE, alpha=0.1) +
  geom_smooth(aes(group=cluster), method = "gam", se = FALSE) +
  scale_colour_manual(values = c("#384B6B", "#E2892B")) + 
  facet_wrap(~ label) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(
    x = "Time ID",
    y = "Response (Mean Cluster Response Bold)",
    color = "Cluster"
  ) +
  theme_Publication()

ggpubr::ggarrange(
  clusterTS + theme(axis.title.x = element_blank()),
  clusterTSTrend,
  ncol = 1,
  labels = c("(A) Timepoint Means", "(B) Spline Trends"),
  common.legend = TRUE,
  legend = "bottom"
)
```

@@

### (Out-of-)Cluster Comparison

@@

To illustrate artifact checks, we added the number of ESM measurements into the comparison and find that the participants in the deterioration cluster (cluster 2) on average completed slightly more ESM surveys in general and reported on more intergroup interactions in particular (see $n$ in [@fig-clusterFeatVarComb] B). This difference in the reported number of interactions might indicate either a clustering artifact or a meaningful difference. The higher average number of interactions in cluster 2 could, for example, indicate a clustering artifact if variances are substantially larger due to the larger samples (e.g., restriction of range in the smaller sample; @kogan2006). In our case, this seems less likely because one out of four variables did not differ in terms of the MAD (i.e., our selected measurement of the time series variance; see plot selection below).

::: scrolling
```{r}
#| label: fig-mad-cluster-comparison
#| message: false
#| warning: false
#| fig-cap: Cluster comparison MAD across variables

@@

# validation features
missingness_ooc <- full_join(
  featData %>%
    mutate(uid = paste(study, PID, sep = "_")) %>%
    group_by(uid, ID) %>%
    summarise(n_feature = n()) %>%
    ungroup(),
  dt_raw %>%
    mutate(uid = paste(study, PID, sep = "_")) %>%
    group_by(uid) %>%
    summarise(n_raw = n()) %>%
    ungroup(),
  by = "uid"
) %>%
  mutate(
    n_rm = n_raw - n_feature
  ) %>%
  filter(
    !is.na(ID)
  ) %>%
  select(ID, n_rm) %>%
  mutate(n_rm_z = scale(n_rm)) %>%
  left_join(data.frame(ID = as.numeric(names(kmeans_final$cluster)),
             cluster = kmeans_final$cluster), by = "ID")
t_missingness <- t.test(
  missingness_ooc$n_rm_z[missingness_ooc$cluster == 1],
  missingness_ooc$n_rm_z[missingness_ooc$cluster == 2],
  na.action = na.omit
)

discrimination_ooc <-
  featData %>% 
  select(ID, EvDayDiscr.post) %>% 
  distinct(ID, EvDayDiscr.post, .keep_all = TRUE) %>%
  mutate(EvDayDiscr.post_z = scale(EvDayDiscr.post, scale = TRUE)) %>%
  replace_na(list(EvDayDiscr.post_z = NA)) %>%
  select(
    ID,
    EvDayDiscr.post_z
  ) %>%
  left_join(data.frame(ID = as.numeric(names(kmeans_final$cluster)),
             cluster = kmeans_final$cluster), by = "ID")

t_discrimination <- t.test(
  discrimination_ooc$EvDayDiscr.post_z[discrimination_ooc$cluster == 1],
  discrimination_ooc$EvDayDiscr.post_z[discrimination_ooc$cluster == 2],
  na.action = na.omit
)

results_df <- rbind(
  results_df,
  data.frame(
    concept = c("EvDayDiscr.post", "missingness"),
    feature = c("mean", "mean"),
    difference = c(t_discrimination$estimate[1] - t_discrimination$estimate[2], t_missingness$estimate[1] - t_missingness$estimate[2]),
    t_value = c(t_discrimination$statistic, t_missingness$statistic),
    df = c(t_discrimination$parameter, t_missingness$parameter),
    p_value = c(t_discrimination$p.value, t_missingness$p.value),
    lwr = c(t_discrimination$conf.int[1], t_missingness$conf.int[1]),
    upr = c(t_discrimination$conf.int[2], t_missingness$conf.int[2])
  )
) %>%
  annotate_significance(., 
                        variableLab = var_meta %>% select(name, label), 
                        join_by = c('concept' = 'name'))

results_df[results_df$concept=="missingness",colnames(results_df)=="label"] <- "Missingness (N removed)"

feature_plot_list <- generate_feature_plots(results_df)
```
:::

```{r}
#| label: render hidden plots
#| echo: false
#| results: asis

for (name in names(feature_plot_list)) {
  # Only hide plots that are not plot_mad
  cat(paste0('<div id="plot_', name, '" style="display: ', ifelse(name == "mad", 'block', 'none'), ';">'))
  print(feature_plot_list[[name]])
  cat('</div>')
}
```

<select id="plot_selector" onchange="showPlot()">
  <option value="plot_mad" selected>MAD</option>
  <option value="plot_median">Median</option>
  <option value="plot_n">Measurements</option>
  <option value="plot_lin">linear slope</option>
  <option value="plot_mac">MAC</option>
  <option value="plot_edf">GAM edf</option>
  <option value="plot_ar01">AR lag1</option>
  <option value="plot_mean">Validation</option>
</select>

<script>
function showPlot() {
  var selectedPlot = document.getElementById("plot_selector").value;
  var plots = document.querySelectorAll('div[id^="plot_"]');

  plots.forEach(function(plot) {
    if (plot.id === selectedPlot) {
      plot.style.display = 'block';
    } else {
      plot.style.display = 'none';
    }
  });
}

// Call showPlot on page load to display the default plot
document.addEventListener('DOMContentLoaded', showPlot);
</script>
